### 2019-11-03 ThesisJo/2_background.tex
2019-11-02

The wave equation has a few solutions of interest, in particular that of a plane and spherical wave:
%
\begin{align}
f_\textrm{plane}(t,\vec{p}_{OS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{OS})} \\
f_\textrm{spheric}(t,\vec{p}_{OS}) &= \frac{A}{\lVert \vec{p}_{OS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{OS} \rVert)},
\end{align}
% 
Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{OS}$ is position of the source $\mathcal{S}$ relative to the observer $\mathcal{O}$. We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to wavelength and array length) and as superimposed spherical waves under near-field conditions.

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers. After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig. \ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at an angle or point at a time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_das.svg}}
\caption{DAS (delay-and-sum) beampattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}%
\label{2_fig_scenario_das}
\end{figure}

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
\caption{MVDR (minimum variance distortionless response) beampattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics.}%
\label{2_fig_scenario_mvdr}
\end{figure}

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_lca.svg}}
\caption{LCA (low complexity adaptive) beampattern is dynamic and discrete (adaptive beamforming). Interference suppression is on-par with MVDR with 15 weight sets to choose from, marginally worse with 6 weight sets.}%
\label{2_fig_scenario_lca}
\end{figure}

Imagine an array $\mathcal{A}$ with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Assume the array channels are digitally delayed to focus at $\udot{S}$, such that the delayed data from the $m$th channel at instant $n$ can be expressed as $x_m[n-\Delta_m, \vec{p}_{mS}]\in\mathbb{C}$. Focus at source $\udot{S}$ requires delays akin to
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$. Subtracting the minimum distance reduces latency and is optional. By making the dependence on $\vec{p}$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$. By definition, the beamformer output $z[n]\in\mathbb{C}$ is a weighted sum of delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights for an array steered to broadside are characterized by the beampattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_m}.\label{2_eq_beampattern}
\end{align}
%
%
%
%
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
The beamformer is called delay-and-sum (DAS) or conventional if the weights are static. A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield. One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}.

@@

Element-space (time-domain) MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile (more on this later). It seeks the weight set that minimizes noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Solving this convex optimization problem with Lagrange multipliers yields
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{a}=\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. In the presence of zero-mean random Gaussian noise, these weights yield the maximum likelihood estimate of the signal and the maximum signal-to-interference-and-noise ratio (SINR). However, active systems or passive ones subjected to multipath or jamming, experience highly correlated noise, interference and signal~\cite{Widrow1982}. Limited amount of data, array miscalibration and model inaccuracies makes matters worse, sometimes resulting in performance inferior to DAS~\cite{Li2006robust}.

To make MVDR robust, it must handle avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness. The literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
%
\begin{itemize}
\item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays. We perform subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure. Some studies construct such matrices directly since are simpler to invert [TK87]. $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89]. 
\item \emph{Temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
\item \emph{Diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%
%
These techniques ensure a numerically well conditioned $\eR$, allow reducing the computational complexity and expose enough data parallelism to facilitate GPU-acceleration.

>>>>

The linear homogeneous wave equation has a few solutions of interest, in particular that of a plane and spherical wave. For a field emitted from source position $\udot{S}$ and observed from $m$th sensor position $\udot{m}$, the solutions are
%
\begin{align}
f_\textrm{plane}(t,\vec{p}_{mS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{mS})} \\
f_\textrm{spheric}(t,\vec{p}_{mS}) &= \frac{A}{\lVert \vec{p}_{mS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{mS} \rVert)},
\end{align}
% 
where $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{mS}$ is position vector $\udot{m}$ to $\udot{S}$. We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to wavelength and array length) and as superimposed spherical waves under near-field conditions.

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (Fig. \ref{2_fig_article_relation}). Image reconstruction coherently combine these receiver outputs to focus at an angle or point at a time. Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_das.svg}}
\caption{DAS (delay-and-sum) beam pattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}%
\label{2_fig_scenario_das}
\end{figure}

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
\caption{MVDR (minimum variance distortionless response) beam pattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics and enough unconsumed degrees of freedom (i.e. enough sensors to cancel the prevalent noise sources).}%
\label{2_fig_scenario_mvdr}
\end{figure}

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_lca.svg}}
\caption{LCA (low complexity adaptive) beam pattern is dynamic and discrete (adaptive beamforming). Interference suppression is on-par with MVDR with 15 weight sets to choose from, marginally worse with 6 weight sets.}%
\label{2_fig_scenario_lca}
\end{figure}

Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$. The delays (normalized by sampling frequency $f_s$) coarsely phase-aligns signals from source $\udot{S}$,
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c\,f_s},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$. Subtracting the minimum distance is mathematically optional, but facilitates physical realization and low latency. By assuming implicit dependence on $\vec{p}$ and data pre-delayed to the pixel of interest, we simplify notation to $x_m[n]\in\mathbb{C}$. Beamformer output $z[n]\in\mathbb{C}$ is, by definition, the weighted sum of delayed data samples,
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where the weights are normalized and $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_{Am}},\label{2_eq_beam pattern}
\end{align}
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
where the start position $\udot{A}$ is arbitrary. Note how the beam pattern depends on both angle and frequency of the incoming wave.

The beamformer is called delay-and-sum (DAS) or conventional if the weights are static (temporally non-varying). Various weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. A class of adaptive beamformers target this limitation by adapting the weights to the dynamic impinging wavefield. One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}. We discuss two versions of it termed by

@@

Element-space (time-domain) MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile (more on this later). It computes the weights that minimize noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Solving this convex optimization problem with Lagrange multipliers yields
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{a}=\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. In the presence of zero-mean random Gaussian noise, these weights yield the maximum likelihood estimate of the signal and the maximum signal-to-interference-and-noise ratio (SINR). However, active systems or passive ones subjected to multipath or jamming, experience highly correlated noise, interference and signal~\cite{Widrow1982}. Limited amount of data, array miscalibration and model inaccuracies makes matters worse, sometimes resulting in performance inferior to DAS~\cite{Li2006robust}.

To make MVDR robust, it must handle avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness. The literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
%
\begin{itemize}
\item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays. We achieve this through subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure. Some studies construct Toeplitz-matrices directly since are simpler to invert [TK87]. $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89]. 
\item \emph{Temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
\item \emph{Diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%
%
These techniques ensure a numerically well conditioned $\eR$, allow reducing the computational complexity and expose enough data parallelism to facilitate GPU-acceleration.
